---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Capr

<!-- badges: start -->

[![CRAN
status](https://www.r-pkg.org/badges/version/Capr)](https://CRAN.R-project.org/package=Capr)
[![codecov.io](https://codecov.io/github/OdyOSG/Capr/coverage.svg?branch=main)](https://codecov.io/github/OdyOSG/Capr?branch=main)
[![Build
Status](https://github.com/OdyOSG/Capr/workflows/R-CMD-check/badge.svg)](https://github.com/OdyOsg/Capr/actions?query=workflow%3AR-CMD-check)

<!-- badges: end -->

The goal of Capr is to provide a language for expressing OHDSI Cohort
definitions in R code. OHDSI defines a cohort as "a set of persons who
satisfy one or more inclusion criteria for a duration of time" and
provides a standardized approach for defining them (Circe-be). Capr
exposes the standardized approach to cohort building through a
programmatic interface in R which is particularly helpful when creating
a large number of similar cohorts. Capr version 2 introduces a new user
interface designed for readability with the goal that Capr code being a
human readable description of a cohort while also being executable on an
OMOP Common Data Model.

Learn more about the OHDSI approach to cohort building in the [cohorts
chapter of the Book of
OHDSI.](https://ohdsi.github.io/TheBookOfOhdsi/Cohorts.html)

## Installation

Capr will be on CRAN for now you can install the current development
version of Capr from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("OdyOSG/Capr")
```

## Example

Capr uses many defaults that match the defaults in Atlas. Creating a
simple cohort is a single line of code. As an example we will define a
cohort of new users of diclofenac described in the [Book of
OHDSI.](https://ohdsi.github.io/TheBookOfOhdsi/SuggestedAnswers.html#Cohortsanswers)

-   New users of diclofenac

```{r example1}
library(Capr)

# Define concepts sets with cs()
diclofenac <- cs(descendants(1124300))

ch <- cohort(drugEra(diclofenac))

ch

```

This is a valid OHDSI cohort definition where persons enter at their
first drug era record of diclofenac and exit at the end of their
observation time.

-   Ages 16 or older

-   With at least 365 days of continuous observation prior to exposure

-   With cohort exit defined as discontinuation of exposure (allowing
    for a 30-day gap)

```{r example2}
diclofenac <- cs(descendants(1124300))

ch <- cohort(
  entry = entry(drugEra(diclofenac, age(gte(16))),
                observationWindow = continuousObservation(-365L, 0L)),
  exit = drugExit(diclofenac)
)

ch
```

-   Without prior exposure to any NSAID (Non-Steroidal Anti-Inflammatory
    Drug)

-   Without prior diagnosis of cancer

```{r example3}
diclofenac <- cs(descendants(1124300), name = "diclofenac")
nsaid <- cs(descendants(21603933), name = "nsaid")
cancer <- cs(descendants(443392), name = "cancer")

ch <- cohort(
  entry = entry(drugEra(diclofenac, age(gte(16))),
                observationWindow = continuousObservation(-365L, 0L)),
  attrition = attrition(withAll(
    exactly(0, drug(nsaid), eventStarts(-Inf, 0, index = "startDate")),
    exactly(0, condition(cancer), eventStarts(-Inf, 0, index = "startDate"))
  )),
  exit = drugExit(diclofenac, persistenceWindow = 30)
)

ch
```

## Save cohort as JSON

OHDSI standard cohorts are represented as json files and can be copy and
pasted into Atlas.

```{r}

path <- file.path(tempdir(), "diclofenacCohort.json")

writeCohort(ch, path)

cat(substr(readr::read_file(path), 1, 100))

```

## Fill in missing concept set details

When saving cohorts it is helpful to fill in the concept details. This
requires a live connection to an OMOP CDM database that includes the
vocabularies used in the cohort definition.

```{r, message=FALSE}

con <- DatabaseConnector::connect(Eunomia::getEunomiaConnectionDetails())

diclofenac <- cs(descendants(1124300), name = "diclofenac")

cat(as.json(diclofenac))

```

```{r, warning=FALSE}

diclofenac <- getConceptSetDetails(diclofenac, con, vocabularyDatabaseSchema = "main")
cat(as.json(diclofenac))

```

## Get Parameterized Cohort Generation SQL

```{r}
 sql <- CirceR::cohortExpressionFromJson(as.json(ch)) %>%
    CirceR::buildCohortQuery(options = CirceR::createGenerateOptions(generateStats = TRUE))

 cat(substr(sql, 1, 220))
```

## 

## Create cohort 'templates'

A Capr cohort template is a function that always returns a Capr cohort.
It can accept arguments that can be used to parameterize any part of a
cohort definition. Capr cohort templates are the recommended approach
for building large numbers of similar cohorts in R.

```{r}

# A Capr cohort template is a function that returns a cohort
drugEraTemplate <- function(ingredientConceptId) {
  
  drugConceptSet <- cs(descendants(ingredientConceptId))
  
  cohort(
    entry = entry(drugEra(drugConceptSet, age(gte(16))),
                  observationWindow = continuousObservation(-365L, 0L)),
    exit = drugExit(drugConceptSet, persistenceWindow = 30)
  )
}


library(dplyr, warn.conflicts = FALSE)

# create a cohort for every single ingredient
df <- DBI::dbGetQuery(con, 
                      "Select * from concept where concept_class_id = 'Ingredient'") %>% 
  tibble() %>% 
  select(concept_id, concept_name) %>% 
  mutate(capr_cohort = purrr::map(concept_id, drugEraTemplate)) %>% 
  mutate(cohort_json = purrr::map_chr(capr_cohort, as.json))

df




```

The capr_cohort column of the dataframe is a list of Capr cohort object.
The cohort_json column contains the json specifications for each cohort.

```{r}
DatabaseConnector::disconnect(con)

```
