---
title: "cohortCapr markdown"
author: "Guus @TheHyve"
date: "2024-07-01"
output: html_document
---
## Set-up knitr
```{r knitr, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::purl(
  input = './cohortCapr_md.Rmd',
  output = './cohortCapr.R'
)
```

## Activate renv
```{r activate renv, include=FALSE}
## R environment activation
# Get renv library
if (!require(renv)) {
  install.packages("renv@v1.0.7")
}
# Get project path
path <- dirname(dirname(rstudioapi::getSourceEditorContext()$path))
if (!("renv" %in% list.files(path))) {
  stop("Working directory is not set correctly. Make sure the working directory is set to the project directory.")
}

## Activate renv
# renv::activate(path)
```

## Set-up libraries
Set up the required libraries, and working directory
```{r housekeeping, echo=FALSE}
## housekeeping
# libraries
library(devtools)
library(RSQLite)
library(tibble)
library(DatabaseConnector)
library(CohortGenerator)
library(CirceR)
library(Capr)
```

##Connect to database
```{r connection details Eunomia, eval=TRUE, include=TRUE}
## Eunomia GI bleed data set
connectionDetails <- Eunomia::getEunomiaConnectionDetails()
```

## Load data set
Use dbConnector library to specified DB and retrieve the GI bleed (default) dataset.
Load this data into a dataframe as well for inspection
```{r load data, messages=FALSE, echo=TRUE}
## Load data
# Connect to database
con <- connect(connectionDetails)

# List all tables
tables <- dbListTables(con)

# Exclude sqlite_sequence (contains table information)
tables <- tables[tables != "sqlite_sequence"]

lDataFramesPre <- vector("list", length=length(tables))
names(lDataFramesPre) <- tables

# Create a data.frame for each table
for (i in seq(along=tables)) {
  lDataFramesPre[[i]] <- dbGetQuery(conn=con, statement=paste("SELECT * FROM '", tables[[i]], "'", sep=""))
}

# Disconnect
disconnect(con)

cat("\n\nTABLES:\n  ")
writeLines(tables, sep = "\n  ")
```

## Concept sets
Define a measurement concept set using Capr and include all descendants
```{r concept sets, echo=TRUE}
## Concept sets
source("./conceptSets.R")
# source("./connectVocabularies.R")

# Establish connection
con <- connect(connectionDetails)

conceptSets$conceptSets <- conceptSets$conceptSets %>%
  # Add details for all concepts (excl. descendants)
  lapply(FUN = getConceptSetDetails, con = con, vocabularyDatabaseSchema = "main")

# Disconnect
disconnect(con)
```

## Included and excluded concepts
Retrieve concepts that have been included/excluded (does not capture descendants)
```{r count occurences, echo=TRUE}
## Count occurrences of each concept in data
# Get countOccurrences function
source("./countOccurrences.R")

# Get links between tables and fields as input
source("./table_linked_to_concept_field.R")

# count occurrences of each concept 'x' and print results; cardiac complications as example
cardiacComplicationsCounts <- 
  countOccurrences(
    conceptSets$concepts$cardiacComplications, c("condition_occurrence", "procedure_occurrence"), 
    lDataFramesPre, links
  ) %>% print()
labTestsCounts <- 
  countOccurrences(
    conceptSets$concepts$labTests, c("measurement"), 
    lDataFramesPre, links
  ) %>% print()
```

## Standard and non-standard concepts
```{r Standard non-standard check}
source('./isStandard.R')
nonStandard <- isStandard(
  concept_table_path = "./project/data/vocabs_14-May-2024/CONCEPT.csv",
  data_concepts_path = "./project/data/phems_variable_list/"
  # save_path = "./project/data/phems_variable_list/is_standard/"
  )

nonStandard
```

## Initial event cohort
```{r Cohort definition, eval=FALSE}
## Cohort definition
# Create cohort definition
ch <- cohort(
  entry = entry(
    # enter patients < 70 years old who have had a lab test
    procedure(conceptSets$conceptSets$cardiacSurgery, age(lt(70))),
    observationWindow = continuousObservation(0, 0),
    primaryCriteriaLimit = "All"
  ),
  attrition = attrition(
    withAny(
      # include all cardiac complications
      atLeast(
        x = 1,
        # include all cardiac complications represented by condition concepts
        query = conditionOccurrence(conceptSets$conceptSets$cardiacComplications)
      ),
      atLeast(
        x = 1,
        # include all cardiac complications represented by procedure concepts
        query = procedure(conceptSets$conceptSets$cardiacComplications)
      )
    ),
    withAny(
      atLeast(
        x = 1,
        # include patients who have had cardiac surgery
        query = procedure(conceptSets$conceptSets$cardiacSurgery)
      )
    ),
    withAny(
      atLeast(
        x = 1,
        # include patients who have had laboratory measurements
        query = measurement(conceptSets$conceptSets$labTests)
      )
    )
  ),
  exit = exit(
    endStrategy = observationExit()
  )
)
```

## Write json expressions and sql queries
```{r json and sql}
## Cohort json and sql
# Generate json for cohort
chJson <- ch %>%
  toCirce() %>%
  jsonlite::toJSON(pretty = TRUE, auto_unbox = TRUE) %>%
  as.character()

# Generate cohort sql query
sql <- CirceR::buildCohortQuery(
  expression = CirceR::cohortExpressionFromJson(chJson),
  options = CirceR::createGenerateOptions(generateStats = FALSE)
)
```

Save the cohort and concept set jsons; these can be imported into ATLAS
```{r Save cohort and concept set json}
write(chJson, paste(path, "./json/cohort.json", sep=""))
for (cs in names(conceptSets$conceptSets)) {
  writeConceptSet(
    x = conceptSets$conceptSets[[cs]],
    path = paste("../json/", cs, "_cs.json", sep="")
  )
}
```

```{r Create and generate cohorts}
# Establish connection
con <- connect(connectionDetails)

# Cohorts to create
cohortsToCreate <- tibble::tibble(
  cohortId = 9876,
  cohortName = "cardiac_arrest",
  sql = sql
)

# Cohort tables
cohortTableNames <- CohortGenerator::getCohortTableNames(cohortTable = "cardiac_arrest")
CohortGenerator::createCohortTables(
  connectionDetails = connectionDetails,
  cohortDatabaseSchema = "main",
  cohortTableNames = cohortTableNames,
)

# Generate the cohorts
cohortsGenerated <- CohortGenerator::generateCohortSet(
  connectionDetails = connectionDetails,
  cdmDatabaseSchema = "main",
  cohortDatabaseSchema = "main",
  cohortTableNames = cohortTableNames,
  cohortDefinitionSet = cohortsToCreate
)

# Get cohort counts
cohortCounts <- CohortGenerator::getCohortCounts(
  connectionDetails = connectionDetails,
  cohortDatabaseSchema = "main",
  cohortTable = cohortTableNames$cohortTable
)

# Disconnect
disconnect(con)


cohortCounts
```

## Get updated DB after querying cohorts
```{r retrieve updated DB}
# Establish connection
con <- connect(connectionDetails)

# List all tables
tables <- dbListTables(con)

# Exclude sqlite_sequence (contains table information)
tables <- tables[tables != "sqlite_sequence"]

lDataFramesPost <- vector("list", length=length(tables))
names(lDataFramesPost) <- tables

# Create a data.frame for each table
for (i in seq(along=tables)) {
  lDataFramesPost[[i]] <- dbGetQuery(conn=con, statement=paste("SELECT * FROM '", tables[[i]], "'", sep=""))
}

# Disconnect
disconnect(con)


cat("Number of persons in dataset: ", length(unique(lDataFramesPost$person$person_id)), 
    "\nNumber of persons in cohort:  ", length(unique(lDataFramesPost$cardiac_arrest$subject_id)))
```

## Filter loaded database for cohort
```{r cohort filter}
cohort_df <- list()
cohort_person_ids <- lDataFramesPost$cardiac_arrest$subject_id
for (table in names(lDataFramesPost)) {
  if ("person_id" %in% names(lDataFramesPost[[table]])) {
    cohort_df[[table]] <- 
      lDataFramesPost[[table]] %>%
      dplyr::filter(person_id %in% cohort_person_ids)
  } else {
    cohort_df[[table]] <- lDataFramesPost[[table]]
  }
}
```

```{r Create new dataset with 1k samples, eval=FALSE, include=FALSE}
new_data <- list()
cohort_person_ids_plus <- append(cohort_person_ids,
  # draw a random sample to make 1000 people including the cohort subjects
  sample(
    unique(lDataFramesPost$person$person_id),
    # sample size = 1k
    size = 1000 - length(cohort_person_ids)
  )
)
for (table in names(lDataFramesPost)) {
  if ("person_id" %in% names(lDataFramesPost[[table]])) {
    new_data[[table]] <- 
      lDataFramesPost[[table]] %>%
      dplyr::filter(person_id %in% cohort_person_ids_plus)
  } else {
    new_data[[table]] <- lDataFramesPost[[table]]
  }
}
```
