---
title: "Searching the OMOP Vocabulary with Capr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Searching the OMOP Vocabulary with Capr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(Capr)
```

Capr provides a set of vocabulary search functions that query the OMOP
concept tables directly. They require a live connection to a database that
contains the OMOP vocabulary schema (concept, concept_ancestor,
concept_relationship, concept_synonym).

All functions accept a `DatabaseConnector` connection and a
`vocabularyDatabaseSchema` argument, and return tidy tibbles with lowercase
column names.

## Connecting to a vocabulary

Any OMOP CDM connection works. The examples below use
[Eunomia](https://github.com/OHDSI/Eunomia) as a small local SQLite
vocabulary for illustration.

```{r connect}
connection <- DatabaseConnector::connect(
  Eunomia::getEunomiaConnectionDetails()
)
vocabularyDatabaseSchema <- "main"
```

For a full vocabulary you can point at a live CDM or a local SQLite file
containing an Athena vocabulary download:

```{r connect-full, eval=FALSE}
connectionDetails <- DatabaseConnector::createConnectionDetails(
  dbms = "postgresql",
  server = "your-server/cdm",
  user = "user",
  password = "password"
)
connection <- DatabaseConnector::connect(connectionDetails)
vocabularyDatabaseSchema <- "cdm_vocab"
```

---

## `searchConcepts()` — keyword search

`searchConcepts()` performs a case-insensitive partial match against
`concept_name`. It is the simplest and most universally compatible search.

```{r search-basic}
# Find standard condition concepts matching "atrial fibrillation"
results <- searchConcepts(
  "atrial fibrillation",
  connection,
  vocabularyDatabaseSchema = vocabularyDatabaseSchema
)
results
```

The returned tibble has columns: `concept_id`, `concept_name`, `domain_id`,
`vocabulary_id`, `concept_class_id`, `standard_concept`, `concept_code`.

### Filtering results

Use `domain` to restrict by OMOP domain and `standardOnly` to include
non-standard concepts:

```{r search-filtered}
# Only drug concepts, include non-standard (source codes)
searchConcepts(
  "metformin",
  connection,
  vocabularyDatabaseSchema = vocabularyDatabaseSchema,
  domain = "Drug",
  standardOnly = FALSE,
  limit = 20L
)
```

---

## `rankedSearchConcepts()` — ranked search

`rankedSearchConcepts()` uses a two-phase approach: a fast ILIKE scan to
identify candidates, then similarity-based ranking. It also scores against
synonyms and counts available source-to-standard mappings.

The ranking algorithm is automatically selected based on the connected DBMS:

| DBMS | Similarity function |
|---|---|
| PostgreSQL | `pg_trgm similarity()` (requires the extension) |
| Snowflake | `JAROWINKLER_SIMILARITY()` |
| Spark / Databricks | normalized `levenshtein()` |
| All others | positional boost (exact > prefix > contains) |

```{r fast-search}
results <- rankedSearchConcepts(
  "atrial fibrillation",
  connection,
  vocabularyDatabaseSchema = vocabularyDatabaseSchema,
  limit = 20L
)
# Results include relevance score and mapping_count
results[, c("concept_id", "concept_name", "relevance", "mapping_count")]
```

The `relevance` column ranges from 0 to ~1.5 (higher is better). Use
`offset` for pagination:

```{r fast-search-page}
# Second page
rankedSearchConcepts(
  "atrial fibrillation",
  connection,
  vocabularyDatabaseSchema = vocabularyDatabaseSchema,
  limit = 20L,
  offset = 0L
)
```

For PostgreSQL users: add a GIN trigram index for dramatically faster search:

```sql
CREATE EXTENSION IF NOT EXISTS pg_trgm;
CREATE INDEX ON cdm_vocab.concept USING GIN (concept_name gin_trgm_ops);
```

---

## `getConceptDescendants()` — concept hierarchy

`getConceptDescendants()` traverses `concept_ancestor` to return the full
descendant tree of one or more seed concepts. This mirrors the
`descendants()` modifier in `cs()`, but returns the full concept detail
tibble so you can review what would be included before building a concept set.

```{r descendants}
# All descendants of "Atrial Fibrillation" (313217)
afib_desc <- getConceptDescendants(
  313217L,
  connection,
  vocabularyDatabaseSchema = vocabularyDatabaseSchema
)
afib_desc
```

The `min_levels_of_separation` column shows distance from the seed. Use
`minLevels` and `maxLevels` to control depth:

```{r descendants-depth}
# Direct children only (level 1)
getConceptDescendants(
  313217L,
  connection,
  vocabularyDatabaseSchema = vocabularyDatabaseSchema,
  minLevels = 1L,
  maxLevels = 1L
)
```

---

## `mapSourceToStandard()` — source code lookup

`mapSourceToStandard()` maps non-standard source codes (ICD-9, ICD-10, NDC,
etc.) to their standard OMOP concepts via the `'Maps to'` relationship.

```{r map-source}
# Map ICD-10 atrial fibrillation codes to standard concepts
mapSourceToStandard(
  c("I48", "I48.0", "I48.1"),
  connection,
  vocabularyDatabaseSchema = vocabularyDatabaseSchema,
  vocabularyId = "ICD10CM"
)
```

Omit `vocabularyId` to search across all vocabularies.

---

## `getConceptInfo()` — lookup by ID

When you already have concept IDs and want the full concept table details:

```{r concept-info}
getConceptInfo(
  c(313217L, 320128L, 201826L),
  connection,
  vocabularyDatabaseSchema = vocabularyDatabaseSchema
)
```

---

## End-to-end: search → build concept set → build cohort

The typical workflow is to use `searchConcepts()` or `rankedSearchConcepts()`
to identify concept IDs interactively, then use those IDs in `cs()` to build
a concept set for a cohort definition.

```{r workflow}
# 1. Find the right concept ID
results <- rankedSearchConcepts(
  "atrial fibrillation",
  connection,
  vocabularyDatabaseSchema = vocabularyDatabaseSchema,
  domain = "Condition"
)
# Inspect results - concept 313217 is "Atrial fibrillation"

# 2. Build a concept set with all descendants
afib_cs <- cs(descendants(313217L), name = "Atrial Fibrillation")

# 3. Optionally fill in display names for Atlas
afib_cs <- getConceptSetDetails(
  afib_cs,
  connection,
  vocabularyDatabaseSchema = vocabularyDatabaseSchema
)

# 4. Build the cohort
afib_cohort <- cohort(
  entry = entry(
    conditionOccurrence(afib_cs),
    primaryCriteriaLimit = "First"
  ),
  exit = exit(endStrategy = observationExit())
)

# 5. Serialize to JSON
toCohortJson(afib_cohort)
```

```{r disconnect}
DatabaseConnector::disconnect(connection)
```
