---
title: "Decompiling Atlas JSON with jsonToCapr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Decompiling Atlas JSON with jsonToCapr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette shows how to convert an Atlas/Circe cohort JSON file into executable
Capr R code using `jsonToCapr()` and `jsonToCaprFile()`.

The common workflow is:

1. Export a cohort definition JSON from Atlas.
2. Convert JSON to Capr code.
3. Source the generated R code.
4. Compile the resulting Capr cohort and (optionally) generate SQL with CirceR.

## Prerequisites

```{r}
library(Capr)
```

You need a cohort JSON exported from Atlas. In this vignette, we refer to it as
`cohort.json`.

## Convert JSON to Capr code in memory

Use `jsonToCapr()` when you want the generated code as a character string.

```{r}
jsonPath <- "cohort.json"

caprCode <- jsonToCapr(jsonPath, mode = "strict")
cat(caprCode)
```

`mode = "strict"` stops when unsupported features are encountered. This is the
safest mode when you need a faithful conversion.

## Write Capr code to an R script

Use `jsonToCaprFile()` to write generated code directly to disk.

```{r}
outRPath <- "cohort_from_atlas.R"

jsonToCaprFile(
  jsonPath = jsonPath,
  outRPath = outRPath,
  mode = "strict"
)
```

The generated script includes `library(Capr)` and can be sourced directly.

## Source generated code and compile

By convention, generated scripts define a `cohortDef` object.

```{r}
source(outRPath)

# cohortDef should now exist in your environment
cohortJson <- compile(cohortDef)
```

At this point, `cohortJson` can be:

- saved and imported back into Atlas, or
- passed to CirceR to generate SQL.

## Optional: validate with CirceR SQL generation

```{r}
sql <- CirceR::buildCohortQuery(
  expression = CirceR::cohortExpressionFromJson(cohortJson),
  options = CirceR::createGenerateOptions(generateStats = FALSE)
)

nchar(sql) > 0
```

If SQL generation succeeds, the converted cohort is structurally valid for CirceR.

## Using skip mode for partial conversion

`mode = "skip"` attempts to continue conversion by omitting unsupported parts and
annotating them in output comments.

```{r}
caprCodeSkip <- jsonToCapr(jsonPath, mode = "skip")
cat(caprCodeSkip)
```

This mode is useful for exploratory migration, but you should manually review
and complete skipped sections before production use.

## Capture skipped items programmatically

For reporting or regression checks, use `returnSkipped = TRUE`.

```{r}
result <- jsonToCapr(
  jsonPath = jsonPath,
  mode = "skip",
  returnSkipped = TRUE
)

names(result)
# [1] "lines" "skipped" "emptyGroupWarnings"

result$skipped
```

## Round-trip pattern

A practical migration pattern is:

1. `jsonToCaprFile()` from Atlas JSON.
2. `source()` generated script.
3. `compile()` to JSON.
4. `CirceR::cohortExpressionFromJson()` + `buildCohortQuery()` as a final validation.

This creates a reproducible, version-controlled bridge from Atlas-authored cohorts
to Capr-authored pipelines.

## Troubleshooting

- If strict mode fails, run with `mode = "skip"` to identify unsupported pieces.
- Review generated `# SKIPPED:` comments and complete the cohort manually in Capr.
- Ensure concept sets referenced by the generated code are present and correctly
  represented before final SQL generation.
