---
title: "Using Capr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using-Capr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Capr provides R users with a language for building computable cohort definitions; 
i.e. definitions that can be translated to SQL code and executed on a observational 
health database in the OMOP Common Data Model format.

While some familiarity with the OMOP Common Data Model standard and the Observational 
Health Data Science and Informatics (OHDSI) open source ecosystem is recommended, this 
tutorial assumes very little prior knowledge.

We define a "cohort" as a set of persons who meet a set of inclusion criteria for 
a duration of time. A cohort can be thought of a set of person-time, the time during 
which persons in a database met the cohort definition. 

A cohort is built by identifying a set of potential index dates and then applying 
inclusion criteria to those potential index dates. Finally persons exit a cohort
either at an explicitly specified time (e.g. X days after index) or when they are no
longer under observation.

In the OMOP CDM data elements are standardized so every OMOP CDM database uses the same 
codes to represent identical clinical data elements. Codes can be looked up in
a publicly available vocabulary search tool called [Athena.](https://athena.ohdsi.org/)

Let's jump into some simple cohort definitions with Capr. We can use an example 
database called `Eunomia` to create reproducible code examples.

The condition concept ID for Gastrointestinal hemorrhage is 192671. We can create a concept set with the `cs` function which works similar to the `c` function for creating vectors in R.

```{r}
library(Capr)

GIBleed <- cs(descendants(192671))

GIBleed
```

The GIBleed concept set will include all descendants of the Gastrointestinal hemorrhage concept 192671.


Creating a simple cohort with the index date at the first occurrence of a GI bleed 
condition occurrence is done in a single line of R code. This cohort takes advantage
of many defaults that match the defaults in Atlas.

```{r}
giBleedCohort <- cohort(condition(GIBleed))

giBleedCohort
```

Cohort geneartion is the process of converting a cohort definition to a table with four columns:
cohort_definition_id, subject_id, cohort_start_date, cohort_end_date. This table contains the person IDs
and dates that people are in the cohort.

```{r setup}


connectionDetails <- Eunomia::getEunomiaConnectionDetails()



sql <- CirceR::buildCohortQuery(CirceR::cohortExpressionFromJson(as.json(giBleedCohort)),
                                CirceR::createGenerateOptions(
                                  generateStats = FALSE
                                ))

cohortsToCreate <- tibble::tibble(
  cohortId = 1,
  cohortName = "GI Bleed",
  sql = sql
)


cohortTableNames <- CohortGenerator::getCohortTableNames(cohortTable = "my_cohort_table")
CohortGenerator::createCohortTables(connectionDetails = connectionDetails,
                                                        cohortDatabaseSchema = "main",
                                                        cohortTableNames = cohortTableNames)
# Generate the cohorts
cohortsGenerated <- CohortGenerator::generateCohortSet(connectionDetails = connectionDetails,
                                                       cdmDatabaseSchema = "main",
                                                       cohortDatabaseSchema = "main",
                                                       cohortTableNames = cohortTableNames,
                                                       cohortDefinitionSet = cohortsToCreate)

# Get the cohort counts
cohortCounts <- CohortGenerator::getCohortCounts(connectionDetails = connectionDetails,
                                                 cohortDatabaseSchema = "main",
                                                 cohortTable = cohortTableNames$cohortTable)
print(cohortCounts)




```




